Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-09 sftrends_manu.Rnw
fig schematic gone

% cleveref options
\crefname{table}{Table}{Tables}
\crefname{figure}{Fig.}{Figs.}
\renewcommand{\figurename}{Fig.}

@@

Environmental conditions are affected by multiple processes acting in a hierarchy across space and time \cite{Oneill89,Levin92}. Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.  Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables. Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis (\cref{fig:schematic})\cite{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years. Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.  The \ac{wrtds} model is a moving window regression that fits a unique set of parameters at each observation point in the time series.  A unique set of weights is used for each regression to control the relevance of observations used to fit the model to the observation at the center of the window. The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.  The final vector used to fit the model at each point weights observations more similar to the center of the window with more importance.  The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.  As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.  The original \ac{wrtds} method is described in more detail in Ref.~\citenum{Hirsch10}.  The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method\cite{Beck15}.  The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.  Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme evens.  The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.  All analyses used the WRTDStidal package for R \cite{Beck16b,RDCT16}

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.  Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic})\cite{Beck15}. Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends. A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.  Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midway points in the time series.

@@

Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.  As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.  Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).  Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}. Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in load contributions before/after upgrade,
\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
\item different nitrogen species will have different changes depending on change in load outputs. 
\end{inparaenum}
The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}

@@

Trends as percent change depend on the mean value, lower values will have larger percent changes.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/schematic.pdf}
\caption{Conceptual diagram illustrating use of \ac{wrtds} to decompose trends in observed nitrogen time series and potential forcing factors that can explain model output.  Results from the model are described as annual and seasonal trends, changes in flow-nutrient dynamics for different time periods, and residual variation independent of time, flow, and season. Relationships between environmental factors (climate, local, regional/historical) and nitrogen trends are more easily related to the separate components of the observed time series using results from the model. }
\label{fig:schematic}   
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/delt_map.pdf}
\caption{The \acl{sfe} Delta and monitoring stations used for analysis. The Delta drains the combined watershed from the Sacramento and San Joaquin rivers (bottom left). All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \citealt{IEP13}).}
\label{fig:delt_map}   
\end{figure}

>>>>

% cleveref options
\crefname{table}{Table}{Tables}
\crefname{figure}{Figure}{Figures}
\renewcommand{\figurename}{Figure}

@@

Environmental conditions are affected by multiple processes acting in a hierarchy across space and time \cite{Oneill89,Levin92}. Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.  Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables. Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years. Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.  The \ac{wrtds} model is a moving window regression that fits a unique set of parameters at each observation point in the time series.  A unique set of weights is used for each regression to control the relevance of observations used to fit the model to the observation at the center of the window. The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.  The final vector used to fit the model at each point weights observations more similar to the center of the window with more importance.  The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.  As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.  The original \ac{wrtds} method is described in more detail in Ref.~\citenum{Hirsch10}.  The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method\cite{Beck15}.  The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.  Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme events.  The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.  All analyses used the WRTDStidal package for R \cite{Beck16b,RDCT16}.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.  Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders\cite{Beck15}. Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends. A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.  Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. These annual and seasonal groupings were chosen for continuity with similar comparisons reported in Ref.~\citenum{Jabusch16} and as approximate twenty year midway points in the time series.

@@

Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.  As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.  Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., )\cite{Cornwell14,Novick15}.  Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}. Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in load contributions before/after upgrade,
\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
\item different nitrogen species will have different changes depending on change in load outputs. 
\end{inparaenum}
The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}

@@

Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/delt_map.pdf}
\caption{The \acl{sfe} Delta and monitoring stations used for analysis. The Delta drains the combined watershed from the Sacramento and San Joaquin rivers (bottom left). All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.}
\label{fig:delt_map}   
\end{figure}
